It's Never Too Late to Learn:\\ \emph{Using Online Machine Learning in Distributed Systems}

### 2016-01-24 writeups/msr_workshop/paper.tex
completed and cleaned up

\documentclass[a4paper,twocolumn]{article}

\usepackage[top=1in, bottom=1in, left=1in, right=1in]{geometry}

@@

\title{
  It's Never Too Late To Learn\\
  \emph{Using Online Machine Learning in Distributed Systems}
}
\author{}
\date{}

@@

Building a modern distributed service that guarantees service-level objectives
is hard. Complexity is introduced by the deployment environment and exacerbated
as the service evolves with time to accommodate hardware/environment changes or
support new features.
For example, to build a simple caching layer that guarantees a certain hit rate,
response latency, and throughput, developers need to take into account the
expected workload accessing the cache, the number of machines available to run
it, the network connecting these machines, the storage hierarchy within each
machine, and interference from other applications running on the network and
machines.  A change in any one of these factors could render the service
woefully inefficient.

Traditionally, this complexity is addressed by using highly-skilled developers
to build custom systems that are optimized for specific configurations and
objective functions.  Is there a general, principled, approach for dealing with
complexity in distributed systems? Hand-tuning the design and configuration of a
system does not scale well with increased complexity or changes in software
design, hardware architecture, deployment environment, and service-level
objectives.

We propose to use online learning to cope with this complexity. Specifically, we
view a {\em decision} made by a distributed system in the framework of
contextual online learning with {\em partial feedback} (where the learner only
gets to observe the result of the action it took, rather than all possible
actions). This is a natural fit for distributed systems where: there is plenty
of context surrounding each decision, decisions are made online, and feedback is
received only for decisions that were executed.  Thus, a hand-designed policy in
a distributed systems can be replaced with policy that has been optimized by a
contextual learning system. By supplying rich contexts to this policy,
developers capture the complexity of our environment without having to
understand it. Additionally, by optimizing the policy continuously, the learning
system can swiftly adapt to changes in the environment or service.

@@

Machine learning has certainly been used in the past to optimize system
decisions. Most of this use has been restricted to either offline settings or
expect {\em full feedback}, whereby a supervised learner is trained on data
that has been annotated with correct labels.
Some systems work in an online setting with full feedback. An example of that
is a caching layer that uses a learning system to dynamically switch between
existing (hand-designed) cache eviction policies, by running all policies in
parallel and measuring their performance.
Systems that work in an online setting with partial feedback
tend to be for news or ad recommendation, but examples from systems
infrastructure exist. For example, caching dynamic query content [], tuning
parameters of multicore data structures [], allocating resources to cluster
computing tasks [], and allocating servers to web applications [].  In all of
these systems, the context used to make a decision, and the decision itself, are
{\em local} to a machine. This is achieved by either devising a decentralized
algorithm in the first place, or giving up on global optimality. We are not
aware of any system that makes ``global'' decisions based on truly distributed
architecture or state.

@@

What is it about distributed systems that makes it difficult to apply online
learning at decision points such as request routing, replica placement, cloud
resource allocation, and failure recovery? This proposal identifies fundamental
challenges posed by distributed systems that must be analyzed and addressed in
order to make online learning a standard tool for distributed infrastructure
systems.  These challenges hold even if we assume that the online learning
system is a black-box that delivers optimal decisions with zero latency. In
other words, they are fundamental to the way we design distributed systems.  We
begin by highlighting some of the factors that contribute to these challenges,
and then list the challenges themselves.

\subsection*{Known unknowns}

@@

\ignore{
However, an ML agent is able to make
decisions online based on inputs from the run-time environment.
}
\ignore{
To be clear, ML has been used in some distributed systems. However, it has not
been used in a general way to make online decisions about core distributed
systems implementation details. We advocate that developers should focus on
high-level objectives of their system, and use online ML to make to compute the
specifics (such as policies for request routing, replica placement, or failure
monitoring) at run-time based on observed metrics.
}

@@

As mentioned earlier, we adopt the framework of contextual online learning with
partial feedback.  To separate the systems concerns we are trying to highlight
from machine learning concerns, we abstract the learning system as a black-box
and assume it provides optimal decisions with zero latency. For a given decision
over a discrete action space, the interface to the learning system is as
follows: given a {\em context} that summarizes the state of the world, the
learning system chooses an {\em action} to take, which our system does and later
reports a {\em reward} indicating how good the action was. An example of a
learning system that supports this interface is the Decision
Service~\footnote{http://aka.ms/mwt}. Now consider a load balancer using that
learning system to make its request-routing decisions. The context for each
decision could be load information of each machine in the cluster, the action
could be a specific machine to route the request to, and the reward could be the
latency of processing the request.

Thus we have reduced our concerns to generating an appropriate context,
executing an action (told to us by the learning system), and reporting reward
information.  We believe that fundamental properties of distributed systems make
even these tasks difficult, as described below.

\subsection*{Locality}

Decisions in a distributed system often rely on contextual information that
resides on diverse and physically disparate components. This information may not
be readily available at the point at which the decision is made (e.g., the load
balancer).  As a result, the context supplied to the learning system may be
stale. What are the effects of staleness, and can the learning system cope with
them (e.g., by relying on ''smoothness'' properties of the context)?  Can we
leverage recent hardware or software techniques to reduce or eliminate
staleness?

Another disparity occurs between the place a decision is made (e.g., the load
balancer) and where the resulting action is taken (e.g., the server selected to
process a request). This means that the enforcement of the action cannot be
guaranteed by the decision maker, and indeed it may be altered or reversed.
Contextual learning systems are capable of tolerating downstream logic
that alters the decision, provided certain independence assumptions
aren't violated and rewards are reported accurately. Are there circumstances in
which distributed systems would violate these assumptions enough to affect
performance?

While reward information is also not available at the decision location, this is
the norm in contextual learning settings, where reward information is only
discernible after a delay and often by a separate subsystem. Thus, we do not
expect this to pose a problem. (The faster reward information is collected, the
faster the resulting data point can be used by the learning subsystem.)

\subsection*{Failures}

The disparity in locality above results in a lack of fate sharing between the
distributed system components involved in a decision.  Failures in any of
these components have implications on the inputs to the learning system. For
example, server failures in a load balanced cluster could mean that certain
context information is staler than others or unavailable altogether. How do we
convey this ''lack of information'' to the learning system, and how does it
cope with sporadically incomplete data? Should failures be encoded in the
context somehow?
How do learning
systems cope with data omission or inability to execute decisions due to
failures? Can the ML agent learn from incomplete or potentially corrupt data?
How sensitive should the learner be to transient conditions in the system?

\subsection*{Hierarchy}

Distributed systems are often designed in a hierarchical manner, where
subcomponents monitor and manage others. These components and layers of
indirection simplify reasoning, building, and maintaining the system at
large. However, it is not clear how hierarchy should be handled with respect to
optimizing decisions via contextual learning. In general contextual learning is
composable, in that decisions of subcomponents can be used to influence
higher-level components, downstream components can make decisions influenced by
the decisions of upstream components, and so on.  Should the context for each
decision be limited to the level it is at in the hierarchy, or is it better for
contexts to span multiple (or all) levels? Should decision points roughly mimic
the hierarchy of the system? Consider deep memory hierarchy, for example. One
could imagine that by capturing contextual information across the entire
hierarchy, one could decide at what level to cache a piece of data, instead of
only deciding on the current level and relying on evictions to move between
levels.

\subsection*{Randomization}

Contextual learning uses exploration (controlled randomization) to discover
potentially good strategies. This is the famous explore-exploit trade-off and is
well established to be necessary for proper online optimization. Randomization
can be encapsulated and handled by the learning system, but there are two
situations in which it might emerge to the systems level. First, if multiple
components are making decisions independently, it is important to ensure that
the randomization used by each (e.g. the seed passed to the learning system)
are independent of each other. Second, many distributed systems already use
randomization inherently, such as randomized load balancing or randomized cache
eviction strategies. Can this randomization be leveraged by the learning
system to achieve its exploration goals naturally?

@@

The approach we have outlined above has at least two drawbacks. One is that we
are using machine-learned models to capture the complexity of a distributed
systems decision. This means that the resulting optimal policy will be opaque to
us, making it difficult to discover what contextual information matters, for
example.  We are independently pursuing work in cracking models open to see what
can be learned from them, but in general this is not always possible nor is it
reliable to draw conclusion from things like learned weights. In effect, our
position is to accept this opacity as the cost of offloading the complexity of
making the right decision to the learning system.

The second drawback is that our proposal only replaces hand-designed policies at
existing decision points. In other words, machine learning is used as an
alternative policy within an existing design and architecture. Can machine
learning influence the design/architecture of the system itself? Systems like
the Decision Service advocate the former approach, but it is conceivable that in
the future distributed systems will be co-designed based on learning objectives.

>>>>

\documentclass[letterpaper,twocolumn]{article}
%\documentclass[pdftex,twocolumn,10pt,letterpaper]{article}
\usepackage[top=1in, bottom=1in, left=1in, right=1in]{geometry}

@@

\title{
  It's Never Too Late to Learn:\\
  \emph{Using Online Machine Learning in Distributed Systems}
}
\author{Hussam Abu-Libdeh and Siddhartha Sen \\ \small{Microsoft Research NYC}}
\date{}

@@

Building a modern distributed service that guarantees service-level objectives
is hard. Complexity is introduced by the deployment environment and exacerbated
as the service evolves with time to accommodate hardware/environment changes or
support new features. Some of the unknown variables are known: a caching layer
must cope with changing workload distributions; a load balancer might have stale
load information from the servers. But some of the unknowns are unknown: a
system could be deployed in a multi-tenant environment where co-located
services affect its performance in unexpected ways.

\ignore{
For example, to build a simple caching layer that guarantees a certain hit rate,
response latency, and throughput, developers need to take into account the
expected workload accessing the cache, the number of machines available to run
it, the network connecting these machines, the storage hierarchy within each
machine, and interference from other applications running on the network and
machines.  A change in any one of these factors could render the service
woefully inefficient.
}

Traditionally, this complexity is addressed by using highly-skilled developers
to build custom systems that are optimized for specific configurations and
objective functions.  Is there a general, principled, approach for dealing with
complexity in distributed systems? Hand-tuning the design and configuration of a
system does not scale well with increased complexity or changes in the
environment or service-level objectives.

We propose to use online learning to cope with this complexity. Specifically, we
cast the decisions made by a distributed system in the framework of
{\em contextual online learning with partial feedback}.  This is a natural fit
for many decisions where: there is plenty of context relevant to the decision, the
decision is made online, and feedback is only received for the action that
was taken.  Thus, we can replace a hand-designed policy for the decision with a
policy that has been optimized by a contextual learning system. By supplying
rich contexts to the learner, developers can capture the complexity of the
environment without having to understand it.  By training the learner
continuously, the system can adapt swiftly to changes in the environment
or service.

@@

Online learning has been used in systems before. Some of these systems require
full feedback, which is not practical, but others operate in the partial
feedback setting.  The latter systems tend to be recommendation systems (e.g.,
for personalized news or ads), but some infrastructure systems exist, such as
for caching dynamic query content or allocating resources to
cluster jobs.  In all of these systems, the contextual decisions are made {\em
locally} within a machine, either because the system is not distributed, or it
uses a decentralized algorithm, or it does not aim for global optimality.
\ignore{We are not aware of any system that makes ``global'' decisions based on truly
distributed architecture or state.}

\ignore{
Machine learning has certainly been used in the past to optimize system
decisions. Most of this use has been restricted to either offline settings or
expect {\em full feedback}, whereby a supervised learner is trained on data
that has been annotated with correct labels.
Some systems work in an online setting with full feedback. An example of that
is a caching layer that uses a learning system to dynamically switch between
existing (hand-designed) cache eviction policies, by running all policies in
parallel and measuring their performance.
Systems that work in an online setting with partial feedback
tend to be for news or ad recommendation, but examples from systems
infrastructure exist. For example, caching dynamic query content [], tuning
parameters of multicore data structures [], allocating resources to cluster
computing tasks [], and allocating servers to web applications [].  In all of
these systems, the context used to make a decision, and the decision itself, are
{\em local} to a machine. This is achieved by either devising a decentralized
algorithm in the first place, or giving up on global optimality. We are not
aware of any system that makes ``global'' decisions based on truly distributed
architecture or state.
}

@@

More importantly, prior work does not abstract, analyze, or address the
fundamental challenges imposed by distributed systems when applying online
learning.  The purpose of this proposal is to identify these challenges and
motivate us to start working on them.  We believe these are systems
challenges: that is, they hold even if we assume black-box access to a
learning system that provides optimal decisions in real-time. By overcoming
these challenges, we stand to gain principled mechanisms and tools for
automating intelligence in our distributed infrastructure systems, such as our
cloud.

\ignore{
and points such as request routing, replica placement,
cloud resource allocation, and failure recovery? This proposal identifies fundamental
challenges posed by distributed systems that must be analyzed and addressed in
order to make online learning a standard tool for distributed infrastructure
systems.  These challenges hold even if we assume that the online learning
system is a black-box that delivers optimal decisions with zero latency. In
other words, they are fundamental to the way we design distributed systems.  We
begin by highlighting some of the factors that contribute to these challenges,
and then list the challenges themselves.
}

\ignore{
\subsection*{Known unknowns}

@@

\ignore{
However, an ML agent is able to make
decisions online based on inputs from the run-time environment.
}
\ignore{
To be clear, ML has been used in some distributed systems. However, it has not
been used in a general way to make online decisions about core distributed
systems implementation details. We advocate that developers should focus on
high-level objectives of their system, and use online ML to make to compute the
specifics (such as policies for request routing, replica placement, or failure
monitoring) at run-time based on observed metrics.
}
}

@@

\ignore{We consider classic decisions made by distributed systems where the
choice of action is discrete, such as: where to route a request, how many replicas to
place (and where), or when to raise an alarm during failures.}

For a given decision, we can view the world as having two components: the
distributed system and the learning system, which we treat as a black box.  The
system interacts with the learner as follows: it provides a {\em context}
to the learner summarizing the state of the world, the learner returns 
an {\em action} to take, the system executes the action and reports a {\em
reward} to the learner indicating how good it was.  An example of a
contextual learning system that supports this interface is the Decision
Service\footnote{http://aka.ms/mwt}.  An example of a decision is which
server a load balancer should send a request to: the context could be the load
information of every server, the action is the chosen server, and
the reward could be the latency of the request.

\ignore{
We separate the world into a As mentioned earlier, we
adopt the framework of contextual online learning with partial feedback.  To separate the systems concerns we are trying to highlight
from machine learning concerns, we abstract the learning system as a black-box
and assume it provides optimal decisions with zero latency. For a given decision
over a discrete action space, the interface to the learning system is as
follows: given a {\em context} that summarizes the state of the world, the
learning system chooses an {\em action} to take, which our system does and later
reports a {\em reward} indicating how good the action was. An example of a
learning system that supports this interface is the Decision
Service~\footnote{http://aka.ms/mwt}. Now consider a load balancer using that
learning system to make its request-routing decisions. The context for each
decision could be load information of each machine in the cluster, the action
could be a specific machine to route the request to, and the reward could be the
latency of processing the request.
}

Thus the system is responsible for generating an appropriate context, executing
an action, and reporting the reward.  The following properties of distributed
systems make these tasks challenging (there may be others).

\paragraph{Locality.}
Decisions in a distributed system may rely on information that resides on
diverse and physically disparate components. This information may not be readily
available at the point of decision (e.g., the load balancer).
As a result, the context supplied to the learner may be stale or
incomplete.  How does staleness affect the learner's ability to optimize
decisions? Can we exploit ``smoothness'' in the context to mitigate this effect?
Can we eliminate staleness using hardware or software techniques?

Another disparity may occur between the point of decision and the
point of execution (e.g., the server chosen to handle a request).  This
separation prevents the decider from enforcing the action; indeed, the action
may be changed or ignored.  Contextual learning systems can tolerate
downstream logic that alters the decision, provided rewards are
correctly reported and certain independence assumptions are not
violated.  Are there distributed systems that violate these assumptions?

It is common for reward information to be reported by a separate
component. We do not expect this to pose a problem beyond the fact that delays 
in reward reporting cause delays in learning.

\ignore{typically reported separately by a asynchronously While reward
information is also not available at the decision location, this is the norm in contextual learning
settings, where reward information is only discernible after a delay and often
by a separate subsystem. Thus, we do not expect this to pose a problem. (The faster reward information is collected, the
faster the resulting data point can be used by the learning subsystem.)
}

\paragraph{Failures.}
The absence of locality results in a lack of fate sharing
between the components involved in a decision.  Failures in any of
these components affect the inputs to the learning system.
For example, server failures in the load balancing example could
result in variable staleness or even missing/corrupt information in
the context.  How should information that is known to be missing 
be conveyed to the learner?  Can corrupt data derail the learner's optimization?

Failures can also prevent a component from executing the action, causing the
system to change the action or ignore it.  Can such failures violate the
independence assumptions mentioned above?  In general, how sensitive should
the learner be to transient conditions in the system?

\paragraph{Hierarchy.}
Components in a distributed system often interact with and manage other
components. This hierarchy simplifies building, maintaining, and reasoning
about the system at large. It is not clear how to take hierarchy into
account when optimizing decisions via contextual learning.  Contextual learning
is generally composable, in that decisions of lower-level
components can safely influence those of higher-level components, and those of
upstream components can influence those of downstream components.   Should the
context for each decision be limited to its level in the hierarchy, or should it span
multiple (or all) levels?  Should the decision points roughly mimic the
hierarchy of the system, or should they cut across? Consider the example of a
deep memory hierarchy. One could imagine capturing contextual
information across the entire hierarchy and using it to decide which level to cache an item in, instead of deciding at each level in isolation and relying on evictions to move between levels.

\paragraph*{Randomization.}

Contextual learning balances exploration (a type of controlled randomization)
with exploitation so that it can discover potentially better policies
without compromising current performance.  Exploration can be completely
encapsulated by the learning system and manifested through the actions it
recommends.  However, many distributed systems already use
randomization, such as for cache eviction, load balancing, 
and replica placement.  Can this randomization be leveraged by the learning
system to achieve its exploration goals at no additional cost?

\ignore{First, if multiple components are making decisions independently, it is
important to ensure that the randomization used by each (e.g. the seed passed to
the learning system) are independent of each other.}

@@

The approach we have outlined has at least two drawbacks that merit their own
investigation.  The first is that we are using machine-learned policies that are
generally opaque, in that we cannot easily inspect them to learn things like
which parts of the context really matter.  Cracking open such policies is
certainly an interesting research direction, but our current view is to accept
opacity as the price we pay for offloading complexity to the learner.

The second drawback is that these policies replace hand-designed policies only
at existing decision points.  In other words, machine learning is used as an
alternative within an existing design and architecture, as opposed to 
influencing the design/architecture of the system itself.  It is very
conceivable that in the future distributed systems will be co-designed with
learners.
